Relative parabolic factorizations of Chevalley groups

### 2015-08-26 congrsubgr-boundgen.tex
Added exceptional isomorphisms

\begin{lemma}\label{lemma:srRI1}
Let $\Phi$ be a root system, $I$ be an ideal of $R$ such that $\sr(I)=1$. Then the relative elementary Chevalley group admits Gauss decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \U(\Phi,I)\U^-(\Phi,I)\U(\Phi,I). \]
As a corollary, the width of $\E(\Phi,R,I)$ with respect to generators $z_\alpha(s,\xi)$ of Theorem~\ref{theorem:Tits-Vaserstein} does not exceed $3\left|\Phi^+\right|+2\rk\Phi-1$.
\end{lemma}
\begin{proof}
Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}\in\SL(2,R,I)$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$ we get matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible top-left corner element $a'$.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we obtain relative Gauss decomposition of $A$
\begin{equation}\nonumber
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'),
\end{equation}
where $\varepsilon\in 1+I$ and $y\in I$. Theorem~\ref{thm:Gauss} implies that the Gauss decomposition of length $3$ hold for every relative elementary Chevalley group.

@@

\begin{lemma}\label{lemma:Z1p}
Let $R=\mathbb{Z}[\sfrac{1}{p}]$ and let $I$ be an ideal of $R$.
Under the assumption of GRH the width of $\SL(2,R,I)$ with respect to generators $z_\alpha(s, \xi)$ of Theorem~\ref{theorem:Tits-Vaserstein} does not exceed $6$.
\end{lemma}

\begin{proof}
Clearly, $I$ is a principal ideal generated by some $m\in\mathbb{Z}$ not divisible by $p$.
Let $g$ be an element of $\SL(2,R,I)$. Write
\[ g=\begin{pmatrix}x & y \\ z & w\end{pmatrix},\ \text{for}\ x=p^\alpha a,\ y=p^\beta bm,\ \text{where}\ a,b,\alpha,\beta\in\mathbb{Z},\ p\nmid a,b. \]

@@

%Now it is easy to derive bounded generation property for $\G(\Phi,\mathcal{O}_S,I)$ from the above statements.
\begin{lemma}
Let $\Phi$ be a classical root system of rank $\geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi,\mathcal{O}_S,I)$ has finite width with respect to generators $z_\alpha(s, \xi)$ of Theorem~\ref{theorem:Tits-Vaserstein}.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leq \asr(\mathcal{O}_S) \leq 2$. 
By Theorem~\ref{thm:BassKolster} we can present any element of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators $z_\alpha(s, \xi)$ and one element of $G_0=\G(\rA_1, \mathcal{O}_S, I)=\SL(2,\mathcal{O}_S,I)$.
Consequently, to prove the statement of the lemma it suffices to express every element $g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$ as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

>>>>

\begin{lemma}\label{lemma:srRI1}
Let $\Phi$ be a root system, $I$ be an ideal of $R$ such that $\sr(I)=1$. Then the relative elementary Chevalley group admits Gauss decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \U(\Phi,I)\U^-(\Phi,I)\U(\Phi,I). \]
As a corollary, the width of $\E(\Phi,R,I)$ with respect to the generating set $\mathcal{Z}$ does not exceed $3\left|\Phi^+\right|+2\rk\Phi-1$.
\end{lemma}
\begin{proof}
Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}\in\SL(2,R,I)$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$ we get matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible top-left corner element $a'$.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we obtain relative Gauss decomposition of $A$
\begin{equation}\nonumber
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'),
\end{equation}
where $\varepsilon\in 1+I$ and $y\in I$. Theorem~\ref{thm:Gauss} implies that the Gauss decomposition of length $3$ hold for every relative elementary Chevalley group.

@@

\begin{lemma}\label{lemma:Z1p}
Set $R=\mathbb{Z}[\sfrac{1}{p}]$ and let $I$ be an ideal of $R$.
Under the assumption of GRH the width of $\SL(2,R,I)$ with respect to the generating set $$\mathcal{Z}(\Sigma_1)=X_{12}(I)\cup X_{21}(I) \cup \{z_{21}(s, \xi) \mid s\in I,\ \xi\in R\}$$ does not exceed $6$.
\end{lemma}

\begin{proof}
Clearly, $I$ is a principal ideal generated by some integer $m\in\mathbb{Z}$ not divisible by $p$.
Let $g$ be an element of $\SL(2,R,I)$. Write
\[ g=\begin{pmatrix}x & y \\ z & w\end{pmatrix},\ \text{for}\ x=p^\alpha a,\ y=p^\beta bm,\ \text{where}\ a,b,\alpha,\beta\in\mathbb{Z},\ p\nmid a,b. \]

@@

%Now it is easy to derive bounded generation property for $\G(\Phi,\mathcal{O}_S,I)$ from the above statements.
\begin{lemma}\ref{lemma:width-dedekind}
Let $\Phi$ be an irreducible classical root system of rank $\ell \geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi,\mathcal{O}_S,I)$ has finite width with respect to the generating set $\mathcal{Z}(\Sigma_\ell)$.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leq \asr(\mathcal{O}_S) \leq 2$. 
Inductively applying Theorem~\ref{thm:BassKolster} and using exceptional isomorphisms for small-rank groups we can present any element of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators from $\mathcal{Z}(\Sigma_\ell}$ and one element of $\G(\{\alpha_\ell, -\alpha_\ell\} \mathcal{O}_S, I)\cong\SL(2,\mathcal{O}_S,I)$.
Consequently, to prove the statement of the lemma it suffices to express every element $g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$ as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

### 2015-08-26 paper.tex
Added exceptional isomorphisms

\title [Parabolic factorizations of congruence subgroups ] {Parabolic factorizations and width of congruence subgroups in Chevalley groups}
\keywords {Chevalley groups, relative subgroups, stability for $\K_1$. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms $\mathcal{Z}$.
\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell))\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell))$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}

@@

In the linear and symplectic case the above results can be extended to even larger class of rings (which includes all Dedekind rings, for example).
\begin{thm}\label{thm:SL2width}
Assume that one of the following is true:
\begin{itemize} 
\item $I$ is an ideal of a (not neccessarily commutative) ring $R$ such that $\sr(I)\leq 2$, $\Phi=\rA_\ell$, $l\geq 2$, $N=3|\Phi^+|+\rk(\Phi) - 3$.
\item $I$ is an ideal of a commutative ring $R$ such that $\sr(I)\leq 3$, $\Phi=\rC_\ell$, $l\geq 2$, $N=3|\Phi^+|+2\rk(\Phi) - 4$
\end{itemize}
Then $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its subgroups isomorphic to $\G(\rA_1, R, I)$. \end{thm}

@@

\subsection{Chevalley groups}
Our exposition of Chevalley groups is standard and follows (...).
\textbf{TODO: Write more about Chevalley groups}

In this paper we work with the irreducible representations of Chevalley groups corresponding to the highest weight
$\varpi_1$ (in the case of a classical $\Phi_\ell$) or $\varpi_\ell$ (in the case $\Phi_\ell=\rE_\ell$). 
The former are the natural vector representations of the classical groups acting on the free modules $V=R^n$ of dimension 
$n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$ respectively.
The latter are the microweight representations of dimension $n=27, 56$ for $\ell=6,7$.

@@

It will be convenient to use the standard numbering for the weights of the natural representations of the classical groups (cp.~\cite[§1B]{St78}):
\[\begin{array}{cll}
  1,2,\ldots, \ell+1 & \text{in the case} & \Phi =\rA_\ell, \\
  1,2,\ldots \ell, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots \ell, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}\]
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

>>>>

\title [Parabolic factorizations of congruence subgroups] {Parabolic factorizations and width of congruence subgroups in Chevalley groups}
\keywords {Chevalley groups, relative subgroups, stability for $\K_1$. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms $\mathcal{Z}$.
\begin{thm} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume that $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis one has $W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 6$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk(\Phi)\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell))$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}

@@

In the linear and symplectic case the above results can be extended to even larger class of rings (which e.\,g. includes all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Assume that one of the following is true:
\begin{itemize} 
\item $I$ is an ideal of a (not neccessarily commutative) ring $R$ such that $\sr(I)\leq 2$, $\Phi=\rA_\ell$, $l\geq 2$, $N=3|\Phi^+|+\rk(\Phi) - 3$.
\item $I$ is an ideal of a commutative ring $R$ such that $\sr(I)\leq 3$, $\Phi=\rC_\ell$, $l\geq 2$, $N=3|\Phi^+|+2\rk(\Phi) - 4$
\end{itemize}
Then $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its subgroups isomorphic to $\G(\rA_1, R, I)$. \end{thm}

@@

\subsection{Chevalley groups}
Our exposition of Chevalley groups is standard and follows~\cite{Ta, S, St78, VP, Va86}.
We denote by $\G(\Phi, R)$ the simply-connected split Chevalley group of type $\Phi$ over arbitrary commutative ring $R$. For example, for each of the four classical series $\Phi=\rA_\ell, \rB_\ell, \rC_\ell, \rD_\ell$ the group $\G(\Phi, R)$ equals $\SL(\ell+1, R)$, $\Spin(2\ell+1, R)$, $\Sp(2\ell, R)$, $\Spin(2\ell, R)$, respectively. Notice that from the uniqueness theorem of M.~Demazure (see~\cite[Corollaire~5.2]{SGA3}) it follows that the exceptional isomorphisms of small-rank groups (which are well-known in the theory of groups over fields) remain valid for Chevalley groups over arbitrary commutative rings $R$. In particular there are isomorphisms of groups $\Spin(5, R)\cong \Sp(4, R)$, $\Spin(6, R)\cong\SL(4, R)$ $\SL(2, R)\cong \Sp(2, R)$.

In this paper we work with the irreducible representations of Chevalley groups with the highest weight
$\varpi_1$ (in the case of a classical $\Phi$) or $\varpi_\ell$ (in the case $\Phi=\rE_\ell$, $\ell=6,7$). The former are the natural vector representations of classical groups acting on free modules $V=R^n$ of dimension $n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$ respectively.
The latter are the microweight representations of dimension $n=27, 56$ for $\ell=6,7$.

@@

It will be convenient to use the standard numbering for the weights of the natural representations of classical groups (cp.~\cite[§1B]{St78}):
\[\begin{array}{cll}
  1,2,\ldots, \ell+1 & \text{in the case} & \Phi =\rA_\ell, \\
  1,2,\ldots \ell, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots \ell, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}\]
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.
